Income and Food Price Transmission\footnote{Revised version of MS 2015310 ``Income and Price Transmission''. Submitted to \emph{Agricultural Economics} for 2. round review. I thank two anonymous reviewers for their valuable comments.}

### 2016-02-29 analysis.Rnw
Analysis section of 2. round version of paper.

% !Rnw root = Master.Rnw

\SweaveInput{data_prep.Rnw}
\SweaveInput{regressions.Rnw}

\section{Empirical analysis\label{sec:analysis}}
This section assesses the model's predictions regarding the relationship between income and price transmission. The model in section \ref{sec:model} predicts an inverse-U shaped relationship between income and price transmission but it does not pin down the actual turning point, which is an empirical problem. Our goal in this section is thus to a) establish whether there is a statistically significant quadratic relationship between income or, alternatively, expenditure and food price transmission and b) to determine the income/expenditure level corresponding to the turning point. Section \ref{sec:data} introduces the data and provides some context. Results from the statistical analysis are presented in section \ref{sec:analysis1} and \ref{sec:analysis2}.

\subsection{Data on domestic and international food prices\label{sec:data}}
The domestic price index data analyzed below is sourced from the International Labor Organization (ILO).\footnote{See \href{http://laborsta.ilo.org/}{http://laborsta.ilo.org/}} ILO maintains the most comprehensive database of domestic food price indexes (FPIs) and consumer price indexes (CPIs) which is used to monitor the cost of living and real wages. The ILO food price database is not widely used, for good reasons, despite a broad country coverage and series going back to the 1970s in many cases. A major issue with this data is that the historical series are not recalculated whenever a new base year is introduced. The same applies when other changes are made such as the inclusion or exclusion of certain items such as beverages or tobacco from the index. This means that many of the series have discrete level shifts associated with these changes. Another issue with the data is that some countries report several series. In some cases one series replace another after a period of overlap. Finally, many of the series contain missing values. To ``tidy up'' the raw data I carried out the following operations. First, I restricted the sample to the period 2005:1-2010:12. Secondly, overlapping series were spliced and in cases where multiple complete series were available from the same country, a single one was selected. Thirdly, series with missing values were discarded. Finally, and most critical for the analysis below, level shifts associated with base year changes etc. were identified and removed.\footnote{Concretely I added the change in the break point from the previous value to the break point and all points succeeding it. The value of the adjusted series at the break point is thus the same as the one preceding it.} In total 140 countries have complete price series for the 2005-10 period considered. An important last thing to know about this data is that most of the low income countries' price indexes derive from prices collected from one or two major markets in the capital. Weights and selected items are derived from household expenditure surveys. Differences in consumption patterns around the world of course translate into differences in weights and in the composition of the food baskets themselves. This is probably the largest drawback of using index data in a cross section analysis of price transmission. That being said, even domestic prices of individual food commodities such as wheat or rice would not refer to the same good due to quality and varietal differences, so it is not the case that focusing on a few selected food items consumed by all countries would solve the problem.

There has been two recent international price spikes; one in 2007-8 and the other in 2010-11. Between January 2007 and June 2008 FAO's food and grain price indexes, which are normalized with respect to the 2002-4 average, rose by 91 and 127 points to index 226 and 268, respectively, corresponding to relative price increases of 68 and 90 percent. From the first peak in mid-2008 and to their their respective troughs in February and September 2009 the two indexes fell by 82 and 113 points before rising back up to index 240 and 261 in February and April 2011, respectively.

The boxplot in figure \ref{fig:infl} depicts the empirical distribution of the sample countries' headline (CPI) inflation, their food (FPI) inflation as well as the change in their FPI relative to the CPI, termed real FPI growth, over the two time periods 2007:1-2008:6 and 2010:1-2010:12.

@@

As mentioned in section \ref{sec:model}, the nominal food price level is not only driven by exogenous world market shocks, it also includes a stochastic trend which it shares with the non-food price level, representing underlying general domestic inflation. By dividing the FPI with the CPI, we obtain an FPI which is purged of headline inflation. The median country experienced an annualized rate of food and headline inflation of \Sexpr{round(median(infl$foodinfl78),1)} and \Sexpr{round(median(infl$cpiinfl78),1)} percent, respectively, in 2007-8. Real FPI growth, which is approximately the difference between the two, was \Sexpr{round(median(infl$rfpigr78),1)} percent for the median country in that period. Price increases were generally lower in the second period. The median country experienced food and headline inflation rates \Sexpr{round(median(infl$foodinfl1011),1)} and \Sexpr{round(median(infl$cpiinfl1011),1)} percent implying a modest real FPI growth of \Sexpr{round(median(infl$rfpigr1011),1)} percent. In both periods the variation in inflation rates is substantial. The standard deviation and interquartile range of 2007-8 food inflation is \Sexpr{round(sd(infl$foodinfl78),1)} and \Sexpr{round(IQR(infl$foodinfl78),1)} percentage points, respectively, versus \Sexpr{round(sd(infl$foodinfl1011),1)} and \Sexpr{round(IQR(infl$foodinfl1011),1)} percentage points in 2010-11. Although median real FPI growth is lower in the second period the variation is similar in both periods (standard deviations of \Sexpr{round(sd(infl$rfpigr78),1)} and \Sexpr{round(sd(infl$rfpigr1011),1)} and interquartile ranges of \Sexpr{round(IQR(infl$rfpigr78),1)} and \Sexpr{round(IQR(infl$rfpigr1011),1)}, respectively). The three remaining panels contain alternative representations the same data. What they show us that there is some systematic variation in the food prices to be explained. As the positively sloped lowess lines indicate, a country hit by a large nominal and real FPI shock in 2007-8 was more likely to be hit by a relatively large shock again in 2010-11 and vice versa. This suggests that there are indeed characteristics which influences a country's response to a world market shock, an assertion that will be explored further in section \ref{sec:analysis1} and \ref{sec:analysis2}.

\subsection{Determinants of domestic food price changes\label{sec:analysis1}}

Table \ref{tab:2007-8-regressions} and \ref{tab:2010-11-regressions} delve deeper into the  relationship between income and domestic food price changes at the time of the 2007-08 and 2010-11 international price spikes, respectively. Each of the columns refer to a regression of FPI inflation or real FPI growth on a set of income related predictor variables. The models can all be written
\begin{equation}
\pi_{i}^{dom}=\delta_{0} + \delta_{1}x_{i} + \delta_{2}x_{i}^{2} + \delta_{3}z_{i} + u_{i}\label{eq:infl_model}
\end{equation}
where $\pi_{i}^{dom}$ refers to either country $i$'s FPI inflation or its real FPI growth in either the 18 month period 2007:1-2008:6 or the 12 month period 2010:1-2010:12, $x_{i}$ refers to a subset of the explanatory variables considered, $z_{i}$ is a vector of control variables and $u_{i}$ is the error term. The covariates, $x_{i}$, include the log of country $i$'s (PPP adjusted) per capita GDP [$\log(GDP)$], its consumption [$\log(cons)$] and food expenditure level [$\log(food)$]. The control variables in $z_{i}$ are country $i$'s percentage dietary energy supply derived from cereals, roots and tubers ($cshare$), its cereal import dependency ratio ($cdep$) and their interaction.\footnote{Cereal dietary energy supply and cereal dependency data are from FAOSTAT and refer to the 2004-06 average. Data on GDP and expenditure is sourced from the World Bank's website. It is the International Comparison Project (ICP) which is behind the expenditure estimates which refer to year 2005.} Summary statistics on these variables are provided in table \ref{tab:sumstat}. Of primarily interest is the quadratic relationship between domestic price changes $\pi_{i}^{dom}$ and expenditure, $x_{i}$. The first 12 rows of table \ref{tab:tp} summarize information about the turning points based on different estimated versions of model \ref{eq:infl_model}.

\begin{center}
[Table \ref{tab:sumstat} around here]
\par\end{center}

Countries that are largely dependent on grain imports to cover their domestic consumption needs cannot use export restrictions as way of stabilizing domestic prices. Because of this vulnerability to international food price fluctuations we would expect to find stronger domestic price increases during international food price spikes among countries with large import dependency ratios. The analysis in section \ref{sec:model} suggests that international price shocks are transmitted the strongest to middle income countries so we would expect food price inflation in these two periods to follow an inverse-U shaped relationship with per capita income and consumption expenditure. Poor households and populations also spend a larger share of their food budget on staples which makes them more vulnerable to price spikes on grains and other internationally traded agricultural commodities. Therefore, if we did account for this, the estimated relationship between income and domestic food price changes could be spurious. By including the $cshare$ variable in the regression this omitted variable bias should be reduced.

\begin{center}
[Table \ref{tab:2007-8-regressions} around here]
\par\end{center}

<<echo=FALSE>>=
<<dom_infl_0708>>
tp3 = -coef(m3)["log(food)"]/(2*coef(m3)["I(log(food)^2)"])
tp6 = -coef(m6)["log(food)"]/(2*coef(m6)["I(log(food)^2)"])
tp4 = -coef(m4)["log(GDP)"]/(2*coef(m4)["I(log(GDP)^2)"])
tp5 = -coef(m5)["log(cons)"]/(2*coef(m5)["I(log(cons)^2)"])
@

Table \ref{tab:2007-8-regressions} considers domestic food price changes during the 2007:1-2008:6 period for the countries in the sample for which there is data on income, expenditure, dietary energy supply and cereal import dependency. Column (1) relates a country's annualized FPI inflation rate in that period to its per capita GDP level. What we see is that average food inflation during the 2007-8 food crisis is increasing in a country's per capita income level up until a certain point after which it starts to decrease. The same applies for consumption and food expenditure in column (2) and (3), respectively. Only the latter set of coefficients are statistically significant though. Specifically, column (3) suggests that among countries with a 2005 per capita food expenditure level of \Sexpr{round(exp(tp3),1)} {[}$\exp(\Sexpr{round(tp3,2)})${]} PPP adjusted dollars or less there is a positive estimated relationship between expected FPI inflation and food expenditure. After that point the predicted marginal effect turns negative. \Sexpr{nrow(data_0708[which(data_0708$food< exp(tp3) & data_0708$food >0), ])} out of the \Sexpr{nrow(data_0708[which(data_0708$food>0), ])} countries in the sample with data on per capita expenditure levels spend less than that on food.

Column (6) focuses on the relationship between average food expenditure and expected real FPI growth in the 2007-8 period. Again there is a significant quadratic relationship between the two and it is remarkably similar to the specification based on FPI inflation in column (3). For countries with per capita food expenditures less than \Sexpr{round(exp(tp6),1)} PPP adjusted dollars (\Sexpr{nrow(data_0708[which(data_0708$food< exp(tp6) & data_0708$food >0), ])}/\Sexpr{nrow(data_0708[which(data_0708$food>0), ])} countries) there is a positive estimated relationship between food expenditure and real FPI growth. For countries with food expenditures higher than that, the relationship becomes negative.

<<echo=FALSE>>=
<<data_0708>>
gdpsort = data_0708[order(data_0708$GDP), ]
@

The GDP and consumption coefficients in column (4) and (5) are also statistically significant and the conclusion from these models is similar to one based on (3) and (6). The GDP model in column (4) predicts that for countries with a per capita GDP below \Sexpr{round(exp(tp4),1)} PPP adjusted dollars (\Sexpr{nrow(data_0708[which(data_0708$GDP< exp(tp4) & data_0708$GDP >0), ])}/\Sexpr{nrow(data_0708[which(data_0708$GDP>0), ])} countries) there is a positive relationship between real FPI increases and income and vice versa. The consumption expenditure model (5) predicts a turning point of \Sexpr{round(exp(tp5),1)} PPP adjusted dollars where \Sexpr{nrow(data_0708[which(data_0708$cons< exp(tp5) & data_0708$cons >0), ])}/\Sexpr{nrow(data_0708[which(data_0708$cons>0), ])} countries have lower average consumption than that. Number \Sexpr{nrow(data_0708[which(data_0708$GDP< exp(tp4) & data_0708$GDP >0), ])} country in the sample in terms of GDP is Nicaragua. Countries in the lower middle income group  therefore generally experienced the largest food increases during the 2007-8 food crisis, whereas low and high income countries were affected less, once we control for differences in consumption and trade patterns.

Countries with higher cereal consumption shares and import dependency ratios experienced larger food price increases as expected although $cshare$ and $cdep$ are not individually significant in most specifications. The negative coefficient on the interaction term means that the effect of $cdep$ on food price growth is decreasing in $cshare$ and vice-versa, but this mitigating effect is negligible.

\begin{center}
[Table \ref{tab:2010-11-regressions} around here]
\par\end{center}

Table \ref{tab:2010-11-regressions} summarizes the results from a set of regressions based on the same explanatory variables as those from table \ref{tab:2007-8-regressions}. The difference is that the dependent variables are domestic FPI inflation and real FPI growth during the 2010:1-2010:12 period. As can be seen, the results are similar to those in table \ref{tab:2007-8-regressions}, cf. also table \ref{tab:tp} .

\subsection{Price transmission\label{sec:analysis2}}
Although the results in section \ref{sec:analysis1} provide insights into the cross sectional variation in domestic price changes during the two recent international food price spikes they have at least two shortcomings. First, none of the estimated relationships represent price transmission as such. Secondly, the estimates are based on two specific time periods, which means that we ignore much the information contained in the individual price series. To accommodate these two shortcomings I estimate country specific price transmission coefficients and regress these on the same income and expenditure predictors as in section \ref{sec:analysis1}.

<<echo=FALSE>>=
<<ts_pt>>
tp1 = -coef(m1)["log(GDP)"]/(2*coef(m1)["I(log(GDP)^2)"])
tp2 = -coef(m2)["log(cons)"]/(2*coef(m2)["I(log(cons)^2)"])
tp3 = -coef(m3)["log(food)"]/(2*coef(m3)["I(log(food)^2)"])
tp4 = -coef(m4)["log(GDP)"]/(2*coef(m4)["I(log(GDP)^2)"])
tp5 = -coef(m5)["log(cons)"]/(2*coef(m5)["I(log(cons)^2)"])
tp6 = -coef(m6)["log(food)"]/(2*coef(m6)["I(log(food)^2)"])
@

Specifically, for each country, I regress its monthly domestic food price changes on its lagged domestic food price changes as well as on contemporaneous and lagged international food price changes. That is, as in the the \cite{IMF2011} study, I estimate country specific autoregressive distributive lag (ADL) models of the type
\begin{equation}
\pi_{t}^{dom}=\delta_{0}+\sum_{s=1}^{S_{y}}\rho_{s}\pi_{t-s}^{dom}+\sum_{s=0}^{S_{f}}\lambda_{s}\pi_{t-s}^{int}+\epsilon_{t}\label{eq:time_series_model}
\end{equation}
where $\pi_{t}^{dom}=\Delta p_{t}$ represents monthly FPI inflation or real FPI growth at time $t$ (both approximated by the first differences to the log of FPI or real FPI) and $\pi_{t}^{int}$ is the FAO food price index at time $t$, also in log differences. Using first differences instead of price levels is a way to avoid the spurious regression problem associated with trending variables. In order to keep things simple, maximum lag lengths are uniformly set to $S_{y}=S_{f}=6$.

Next I calculate the long run effects of changes in the the FAO food price index on domestic food inflation and real FPI growth, the first of which is a standard measure of food price transmission. The long run multiplier is given by 
\begin{equation}
\theta=\frac{\sum_{s=0}^{S_{f}}\lambda_{s}}{1-\sum_{s=1}^{S_{y}}\rho_{s}}.
\end{equation}
Finally, I regress the price transmission estimates, $\hat{\theta}$, on the same explanatory variables as in table \ref{tab:2007-8-regressions} and \ref{tab:2010-11-regressions}. That is, I estimate the model
\begin{equation}
\hat{\theta}_{i} =\delta_{0} + \delta_{1}x_{i} + \delta_{2}x_{i}^{2} + \delta_{3}z_{i} + u_{i}\label{eq:ptreg}.
\end{equation}
The results from this exercise are summarized in table \ref{tab:lri-regressions} and \ref{tab:tp}. As table \ref{tab:lri-regressions} shows, we find the same inverse-U shaped estimated relationships between income/expenditure and price transmission as we did with income/expenditure and domestic price changes in table \ref{tab:2007-8-regressions} and \ref{tab:2010-11-regressions}. For the specifications with significant income and expenditure coefficients, i.e. (1)-(5), the estimated turning points again imply that roughly one third of the countries are on the increasing part of the curve, cf. \ref{tab:tp}.

\begin{center}
[Table \ref{tab:lri-regressions} around here]
\par\end{center}

<<echo=False>>=
<<panel_pt>>
tp1 = -m1$coefficients[8]/(2*m1$coefficients[9])
tp2 = -m2$coefficients[8]/(2*m2$coefficients[9])
tp3 = -m3$coefficients[8]/(2*m3$coefficients[9])
tp4 = -m4$coefficients[8]/(2*m4$coefficients[9])
tp5 = -m5$coefficients[8]/(2*m5$coefficients[9])
tp6 = -m6$coefficients[8]/(2*m6$coefficients[9])
@

>>>>

% !Rnw root = master.Rnw

\section{Empirical analysis\label{sec:analysis}}
This section assesses the impact of the determinants discussed above on international-to-domestic aggregate food price transmission. Section \ref{sec:data} introduces the data and provides some context. Results from the statistical analysis are presented in section \ref{sec:analysis1} and \ref{sec:analysis2}.

\subsection{Data\label{sec:data}}

<<echo=FALSE>>=
load("infl_0708_1011.Rda")
load("ts_dat.Rda")
n.full = nrow(table(ts.dat$id))
n.sub = nrow(infl)
infl = infl[infl$iso3c != "BLR", ]
@

\subsubsection{Domestic and international food prices}
Data on domestic prices are sourced from the International Labor Organization (ILO).\footnote{See \citet{ilo16}} ILO maintains a comprehensive database of domestic food price indexes (FPIs) and consumer price indexes (CPIs) that is used to monitor the cost of living and real wages. There are not many studies which draw on this database, however, despite a broad country coverage and long monthly price series. A major issue with this data is that historical series are not recalculated whenever a new base year is introduced. The same applies when other changes are made such as the inclusion or exclusion of certain items such as beverages or tobacco from the index. This means that many of the series have discrete level shifts associated with these changes. In some cases one series replace another after a period of overlap. Finally, many of the series contain missing values. To ``tidy up'' the raw data I carried out the following steps First, I restricted the sample to the period 2005:1-2014:12. Secondly, in cases where multiple complete series were available from the same country, a single one was selected. Thirdly, series with missing values were discarded. Fourthly, and most critical for the analysis below, level shifts associated with base year changes etc. were identified and removed.\footnote{Concretely I added the change in the break point from the previous value to the break point and all points succeeding it. The value of the adjusted series at the break point is thus the same as the one preceding it.} Finally, all the series were normalized with respect to a common base period (Jan 2007). In total \Sexpr{n.full} countries have complete price series for the entire ten year period considered whereas, for example, there are \Sexpr{n.sub} countries with complete price series covering the 2005-11 subperiod. There is therefore a trade-off between the number of complete price series and the length of the series.

An important detail to know about this data is that many of the low income countries' price indexes derive from prices collected from a few major markets in the capital. Furthermore, weights and selected items are derived from household expenditure surveys which often target certain groups such as the urban middle class. Differences in consumption patterns around the world of course translate into differences in weights and in the composition of the food baskets themselves. This is probably the largest drawback of using index data in a cross section analysis of price transmission. That being said, even domestic prices of individual food commodities such as wheat or rice would not refer to the same good due to quality and varietal differences, so it is not the case that focusing on a few selected food items consumed by all countries would necessarily solve the problem.

International food prices are represented by the FAO food price index (FAO FPI) and in some cases its ``real'' version, namely the nominal FAO FPI deflated by the Manufactures Unit Value (MUV) index published by the World Bank.\footnote{see \citet{fao16} and \citet{worldbank16a}.}

\subsubsection{Domestic food inflation during the food crises}

There has been two recent international price spikes; one in 2007-8 and the other in 2010-11. The former spike especially is often referred to as ``the food crisis'' in the literature \citep[e.g.][]{Headey2008, Headey2013, timmer10, Ivanic2012, Martin2011}. Between January 2007 and June 2008 the FAO FPI rose by 91 points to index 226, corresponding to a 68 percent relative price increase. From the first peak in mid-2008 and to the trough in February 2009 the FAO FPI fell by 82 points before rising back up to index 240 in February 2011.

The boxplot in figure \ref{fig:infl} depicts the empirical distribution of the sample countries' headline (CPI) inflation, their food (FPI) inflation as well as the change in their FPI relative to the CPI, termed real FPI (RFPI) growth, over the two time periods 2007:1-2008:6 and 2010:1-2010:12.

@@

The median country experienced an annualized rate of food and headline inflation of \Sexpr{round(median(infl$fpigr.0708),1)} and \Sexpr{round(median(infl$cpigr.0708),1)} percent, respectively, in 2007-08. Nominal food prices are not only driven by exogenous world market shocks; they follow a stochastic trend together with the non-food prices, representing underlying general (core) inflation. Dividing with the CPI is a simple way of purging the FPI of underlying domestic inflation. RFPI growth, which is approximately the difference between the FPI and CPI inflation rate, was \Sexpr{round(median(infl$rfpigr.0708),1)} percent for the median country in 2007-08.\footnote{A problem with the real FPI is that domestic non-food prices are also affected by international commodity shocks. Low or even negative growth in real domestic food prices in these two periods could therefore reflect a response to the concurrent surge in non-food commodities, oil in particular, rather than a puzzling negative response to an international food price shock.} Domestic price increases were generally lower in 2010-11. The median country experienced food and headline inflation rates \Sexpr{round(median(infl$fpigr.1011),1)} and \Sexpr{round(median(infl$cpigr.1011),1)} percent and a modest RFPI growth of \Sexpr{round(median(infl$rfpigr.1011),1)} percent. There is, however, a substantial variation in the inflation rates in both periods.

The three remaining panels of figure \ref{fig:infl} contain alternative representations the same data. There appears to be some systematic variation; as the (mostly) positively sloped lowess lines indicate, a country hit by a large nominal and real FPI shock in 2007-8 was more likely to be hit by a relatively large shock again in 2010-11 and vice versa. This suggests that there are indeed characteristics which influences a country's response to a world market shock, an assertion that will be explored further in section \ref{sec:analysis1} and \ref{sec:analysis2}.

\subsubsection{Price transmission determinants}
Table \ref{tab:cs_data} in the appendix lists the variables that, based on the discussion in section \ref{sec:model}, are considered potentially important determinants of price transmission. A country's PPP adjusted per capita GDP is considered a proxy for its income level. Other covariates considered are a country's percentage dietary energy supply derived from cereals, roots and tubers, its cereal import- and export shares, a dummy indicating whether the country is a landlocked developing country (LLDC), its trade openness defined as the sum of its imports and exports relative to its GDP and a dummy indicating whether there is a grain state trading enterprise (STE) in the country. In addition there are two indexes included representing the regulatory performance of the country in question and the quality of its infrastructure. These two indexes are abbreviated DTF (Distance to Frontier) and LPI (Logistics Performance Index), respectively.\footnote{The covariates are similar to those of \citet{greb12}.}

\subsection{Determinants of domestic food price changes during the food crisis\label{sec:analysis1}}
Table \ref{tab:07-08-fc-regressions} considers the cross sectional variation in domestic food price changes during the 2007-8 food crises. Each of the columns refer to a regression of FPI inflation or real FPI growth on the explanatory variables discussed above as well as the percentage change in a country's domestic currency/US dollar exchange rate during the food crisis. All variables except for GDP and the dummies have been standardized in order to facilitate comparisons of effect sizes.

\begin{center}
[Table \ref{tab:07-08-fc-regressions} around here]
\par\end{center}

Columns (1) and (3) are based on regression models that include the full set of explanatory variables. Income has a significant inverse-U shaped effect as predicted. The other significant covariates are a country's cereal import and export shares, its trade openness, the STE dummy and its LPI. Grain exporters experienced more food inflation on average whereas importers experienced less. The coefficient to the STE dummy is negative and large in magnitude suggesting that countries with grain state trading enterprises were able to control domestic food prices better than other countries. Two of the significant variables have unexpected signs, namely a country's cereal import share and its quality of infrastructure and trade as summarized by its LPI. It is not clear why large grain importers experienced lower food inflation during the food crisis. It could be though, that large food importers intervene more in their food markets than other countries. This interpretation is consistent with the finding by \citet{Lee2013} that large food importers have lower food inflation in general but more research is needed before we can say anything conclusive about this. Better infrastructure should imply more integrated markets and stronger price transmission but countries with a high LPI score experienced lower food inflation on average during the food crisis. The effect on real FPI growth is insignificant though, which would suggest that these countries also experienced lower non-food inflation. Countries with depreciating currencies experienced significantly stronger FPI inflation as expected but the effect on real FPI growth is insignificant. The latter finding is not surprising given that exchange rate changes affect domestic food- as well as non-food prices.

A concern is that some of the variables are highly correlated which leads to large standard errors and insignificant effects. Therefore, the subset of covariates included in columns (2) and (4) are those which minimizes the Akaike information criterion (AIC) as determined by stepwise regression.\footnote{This was carried out using the \emph{MASS} package in \textsf{R} \citep{venables02}} Stepwise regressions have been criticized for leading to models where the remaining terms appear to be more important than they actually are. However, as can be seen, no substantial changes result from this exercise.

What effect did the government interventions have on domestic food inflation during the food crisis? Table \ref{tab:int-regressions} in the appendix reports results from a set of regressions similar to those in table \ref{tab:07-08-fc-regressions} except that a set of dummies are included indicating whether or not the government intervened in various ways during the food crisis.\footnote{These dummies are based on table 1 in \citet{Demeke2009}. ``Export'' = 1 if the country Restricted or banned export, ``Price'' = 1 if the country administered price control or restricted private trade, ``Stock'' = 1 if the country released stocks (public or imported) at subsidized prices'', ``Tariff'' = 1 if the country reduced tariffs or customs fees on imports and ``Tax'' = 1 if the country suspended or reduced VAT or other taxes, all as of 1. December 2008.\label{int_note}} Some of the coefficients to these intervention dummies are substantial but they are all statistically insignificant. Table \ref{tab:1011-regressions} in the appendix contains results related to the 2010-11 international price spike. Again, estimated effects are similar to those in  table \ref{tab:07-08-fc-regressions} but most are statistically insignificant. In fact,  the cereal consumption share is the only robustly significant predictor of domestic food inflation in 2010-11.

\subsection{Determinants of price transmission \label{sec:analysis2}}
Although the results in section \ref{sec:analysis1} provide interesting insights into the cross sectional variation in domestic price changes during the two recent international food price spikes they have at least two shortcomings. First, none of the estimated relationships represent price transmission as such as there may be the other and perhaps more important causes of the observed domestic price changes. Secondly, the estimates are based on a specific short time period, which means that we ignore much the information contained in the individual price series and it also raises questions about the reliability of the results. To accommodate these shortcomings I approach the issue from two different angles. First, in line with the small existing literature on global food price transmission patterns, I estimate a set of long run price transmission rates based on country specific time series models. These estimates are then subsequently regressed on the same country characteristics as in section \ref{sec:analysis1}. Secondly, I estimate the effects of the predictor variables on short run price transmission directly in a panel model.

\subsubsection{Meta analysis of estimated long run price transmission rates\label{sec:meta-analysis}}
This section carries out a meta-regression analysis to see whether a country's estimated long run price transmission rate varies systematically with the proposed determinants of price transmission.

Following the literature, monthly domestic food price changes are regressed on lagged domestic food price changes as well as on contemporaneous and lagged international food price changes and exchange rates \citep[see][]{Ianchovichina2014, IMF2011, Kalkuhl2014}. That is, for each country $i$ an autoregressive distributive lag (ADL) models is fitted
\begin{equation}
\pi_{i,t}^{dom}=\delta_{i,0}+\sum_{s=1}^{p}\rho_{i,s}\pi_{i,t-s}^{dom}+\sum_{s=0}^{q}(\lambda_{i,s}\pi_{t-s}^{int}+\phi_{i,s}er_{t-s})+\epsilon_{i,t}\label{eq:time_series_model}
\end{equation}
where $\pi_{i,t}^{dom}$ represents country $i$'s monthly FPI inflation or RFPI growth at time $t$ (both approximated by the first differences to the log of its FPI or RFPI), $\pi_{t}^{int}$ denotes the FAO FPI or, in case of the RFPI growth model, the FAO RFPI at time $t$ and $er_{i,t}$ is the country $i$'s domestic currency/US\$ exchange rate, both in log differences.

The ADL model is appropriate when the causal impact is unidirectional. This could be the case when the domestic price is from a small country relative to the world market but, more often, this assumption is invoked when the domestic price variable is a Food Price Index (FPI), i.e. the food component of the Consumer Price Index (CPI).

Prior to estimation I tested each of the series for unit roots and cointegration with international prices.\footnote{I tested for stationarity with the KPSS test \citep{kwiatkowski92} and for unit roots with the augmented Dickey-Fuller test \citep{dickey79}. I used the Phillips and Ouliaris ``Pz'' test \citep{phillips90} to test for cointegration. These tests are implemented in \textsf{R} via the packages \emph{tseries} \citep{trapletti15} and \emph{urca} \citep{pfaff08}, respectively.} Results indicated that the domestic FPI and real FPI series, are nonstationary in levels but stationary in first differences. Domestic price levels were also found not to be cointegrated with world prices. Consequently, each model was estimated in growth rates rather than levels in order to avoid spurious regressions.

Next, for each country, I calculated the long run multiplier (LRM) associated with a change in the the FAO food price index given by
\begin{equation}
\theta_{i}=\frac{\sum_{s=0}^{q}\lambda_{i,s}}{1-\sum_{s=1}^{p}\rho_{i,s}}.
\end{equation}

<<echo=FALSE>>=
library(dplyr)
load("lrm_fpi_l1_12.Rda")
fpi = select(lrm.fpi.l.1_12, -iso3c)
load("lrm_rfpi_l1_12.Rda")
rfpi = select(lrm.rfpi.l.1_12, -iso3c)
fivenum.fpi = data.frame(sapply(fpi, summary))
fivenum.rfpi = data.frame(sapply(rfpi, summary))
n = nrow(fpi)
@

The LRM is a common measure of price transmission. \cite{IMF2011}, for example, contains an analysis of food price price transmission based on a dataset of monthly FPI inflation from 31 advanced economies and 47 emerging and developing economies over the period 2000 - 2011. Their median estimated LRM pass-through of a 1 percent increase in international food inflation is 0.18 percent in developed countries and 0.34 percent in emerging and developing countries. Similarly, \citet{Ianchovichina2014} report an average estimated LRM for the eighteen Middle East and North Africa (MENA) countries of 0.25 and the majority of the countries are reported to have  food price pass-through LRMs in the order of 0.2-0.4. Both studies use 12 lags in their country specific ADL models, i.e. they assume $p=q=12$.\footnote{\cite{IMF2011} do not include exchange rates in their regressions.} My own estimates based on a sample of \Sexpr{n} countries covering the period 2005:1 to 2014:12 are very similar. The average and median FPI LRM from a model with 12 lags of all variables are \Sexpr{round(fivenum.fpi["Mean", "V12"], 2)} and \Sexpr{round(fivenum.fpi["Median", "V12"], 2)}, respectively. That being said, it is not obvious that so many lags should be included in the models. When lag length selection is based on the Akaike Information Criterion (AIC), for example, the optimal number lags in most of the models turn out to be $p=q=1$.\footnote{For each country, I calculated the AIC for each of the $12^2$ combinations of $(p,q)\, \in (1,..,12)$. Then I selected the $(p,q)$ combination with the lowest AIC.} But when a common lag length of 1 is used for all countries, average and median FPI transmission LRMs are much lower, namely \Sexpr{round(fivenum.fpi["Mean", "V1"], 2)} and \Sexpr{round(fivenum.fpi["Median", "V1"], 2)}, respectively, whereas the average of the standard errors of the LRM estimates very similar.\footnote{I also tested for residual autocorrelation with the Durbin-Watson test. There was no evidence of residual autocorrelation in either case. Standard errors of the LRMs are based on the so called Bewley transformation of the ADL model instrumented with lags of the dependent variable.} As a compromise between the long lag lengths used in the literature which might cause problems with overfitting and the short optimal lag lengths implied by the AIC criterion which leads to LRM estimates that are lower than commonly found in the literature, the country specific models are based on common lag lengths of $p=q=6$. This choice leads to average long run FPI and RFPI transmission rates of \Sexpr{round(fivenum.fpi["Mean", "V6"], 2)} and \Sexpr{round(fivenum.rfpi["Mean", "V6"], 2)}, respectively. One interpretation of the very low RFPI LRM average is that shocks to a country's FPI is passed on almost fully to its non-FPI in the long run, a phenomenon known as propagation in the literature \citep[e.g.][]{pedersen10}.

The long run transmission estimates, $\hat{\theta}$, are regressed, on the same explanatory variables as in table \ref{tab:07-08-fc-regressions}. However, since the dependent variable is an estimate rather than an observed value and the model now represents an estimated dependent variable (EDV) regression \citep[see][]{hanushek74, lewis05}. A common approach in this situation is to use Weighted Least Squares (WLS) estimation with weights $w_{i}$ that are inversely proportional to the standard error of $\theta_{i}$ \citep{saxonhouse76}. This way more precise estimates are given a higher weight in the regression. The main reason though for using WLS rather than OLS is that the latter may suffer from heteroskedasticity which leads to inefficient coefficient estimates inconsistent standard errors. The inconsistent OLS standard errors can, however, easily be corrected with a robust standard error estimator. \citet{lewis05} uses formal reasoning backed up by a Monte Carlo analysis to argue that OLS with robust standard errors is likely to produce more reliable estimates than WLS based on inverse standard error weights, and that the latter is likely to produce downward biased standard errors.

Table \ref{tab:fpi-pt-regressions} presents estimates of FPI transmission determinants based on OLS [columns (1)-(2)] as well as WLS [columns (3)-(4)], both with robust standard errors. Again, all variables except per capita GDP and the dummies have been standardized and, as in table \ref{tab:fpi-pt-regressions}, column (1) and (3) represent regressions that are based on the full set of covariates whereas columns (2) and (4) represent regressions with the lowest AIC.

\begin{center}
[Table \ref{tab:fpi-pt-regressions} around here]
\par\end{center}

First, as can be seen, results are robust to the choice of estimator. Secondly, although the results in table \ref{tab:fpi-pt-regressions} relate to determinants of FPI transmission rather than FPI inflation as in table \ref{tab:07-08-fc-regressions}, the estimated coefficients and significance patterns are similar across the two sets of regressions. A large grain consumption and grain import share is associated with lower price transmission and so is having a grain state trading enterprise. Large grain exporters, on the other hand, experience larger price transmission on average and so do more open countries. The signs of some of these coefficients could reflect systematic differences in the degree of government intervention in the domestic food market as discussed in section \ref{sec:analysis1}. Income again has a highly significant inverse-U shaped affect of FPI transmission and, oddly, countries with better infrastructure are once more found to experience lower price transmission on average. A possible explanation for the latter effect could be, that the LPI is simply another proxy for income.

Table \ref{tab:rfpi-pt-regressions} in the appendix contain regression results based on estimated long run RFPI transmission rates from country specific time series models with 6 lags. The signs of the coefficients are similar to those in table \ref{tab:fpi-pt-regressions} but most effects are statistically insignificant. Similarly, table \ref{tab:fpi-pt1-regressions} and \ref{tab:rfpi-pt1-regressions} in the appendix contain results based on FPI and real FPI transmission rates calculated from time series models with a single lag. These also point to only a few significant predictors of price transmission. A plausible reason is that the LRM estimates based on models with a single lag are relatively more noisy and therefore more difficult to explain.

\subsubsection{A panel model of short run price transmission determinants}
The regression results in table \ref{tab:07-08-fc-regressions} and \ref{tab:fpi-pt-regressions} are based on a small number of observations which decreases the precision of the estimates. Furthermore, the estimates are potentially subject to unobserved heterogeneity bias. The last set of regressions therefore exploits the panel structure of the dataset to control for time invariant unobserved heterogeneity which potentially confounds the estimated relationships. Specifically the following model is fitted on monthly price data covering the full 2005-15 period 
\begin{equation}
\pi_{it}^{dom}=\beta_{0}+
\sum_{s=1}^{6}\rho_{s}\pi_{it-s}^{dom}+
\sum_{s=0}^{1}(\lambda_{s}\pi_{t-s}^{int}+\phi_{s}er_{it-s})+
\beta_{1}\pi_{t-1}^{int}\cdot x_{i} +
u_{it}\label{eq:panel_model},
\end{equation}
where $x_{i}$ is the vector of country characteristics.
